### 2021-05-05 papers/SYRCoSE/response.txt

We would like to thank the reviewers for their valuable comments. 
We also thank the Reviewer #2 for the reference to the interesting paper relevant to our work.
We proceed with answering the Reviewer #3 questions.

> The Coq proof assistant should be correctly cited, not only particular Coq libraries

We will cite the relevant papers in camera-ready version of the paper.

> Unfortunately, the Event Structure (ES) of Fig2 does not correspond exactly to the set of executions of Fig1: Fig1 suggests that the read operation is carried out after all write actions; this cannot be concluded from the ES of Fig2.

> Fig 9: the prime event structure forgets the difference between the green and the black preorder, therefore it does not express the requirement that in order R(x,2) to occur it must be "immediately preceded" by W(x,2) (in the sense "last write of x before the read"). If you consider this as a prime event structure an execution W(x,2) . W(x,1) . R(x,2) is not forbidden, as W(1,x) and W(2,x) are independent

We answer both of this questions below, as they are connected to each other.

The requirement that the "read is preceded by the last write to the same location" 
in fact corresponds to the sequentially consistent memory model. 
In other models (e.g. total store order TSO, or release/acquire RA) this condition is relaxed
(these models do not require an existance of total order on all events within a single execution (configuration)
and thus in general there is no unique "last" write to read-from).

Our construction of event structure guarantees only the basic condition --- 
acyclicity of program order and reads from. It is, however, a necessary 
precondition for a wide class of memory models, including sequential consistency,
total store order, release/acquire, etc.

Whenever the configuration is extracted from the event structure
it can be checked to additionally satisfy the consistency predicate 
of a particular memory model 
(we briefy mention this in the paper, at the end of Section V).

Moreover, a po+rf acyclic prime event structure together with 
a consistency predicate for a particular memory model 
can be seen as a prime event structure with general (non-binary) conflict. 
In other words, a consistency predicate induces another notion of the conflict.
We are currently working on the theory of prime event structures with 
general conflict that would give us a more solid theoretical background 
for expressing various memory consistency models in the framework of event structures.

As for Fig.1 and Fig.2, you are right here. 
Fig.1. corresponds to sequentially consistent executions of the program,
but the Fig.2 presents po+rf acyclic event structure.
We will try to address this problem and remove the confusion.

> A bit more information about the Coq libraries you use, and why yo use them would be welcome. 
There is space to say a bit more on the approach, may be also for those not "fluent in Coq".
The library is not presented at all, only the application. One would like to get some information about the library itself

We will try to address this problem and give more detailed description
of both our library itself and its dependencies.

> Fig 7: the program example suggests that there is a third behavior of the read thread which runs fully in parallel with the right threads: it should also be possible to "read too early", that is, to observe R(4,\bottom) or get an error ... depending on the semantic choices (the rules of Fig 6 suggest the 1st option, as "errors" are not foreseen)

Yes, you are right. The event structure on Fig.7 is incomplete (i.e. it does not contain the read R(4,\bottom). 
We omit it only for presentation purposes. We will consider to add the missing event to avoid the confusion.

> Fig 8, Store rule: I suppose here e_rf is an error

Yes, it is a typo.

> Theorem 1: did you use Coq to make this proof ? It should be done

Yes, we have a proof in Coq.

### 2021-05-07 papers/SYRCoSE/intro.tex
intro: link to the git repo
minor grammar
intro: minor
*: apply Reviewer #3 suggestions

The same information can be encoded in a single 
event structure containing $6$ events in total
(see \cref{fig:intro-es}). 
In the event structure there are two types of edges 
between the events. The grey arrows $e_1 \arrowCA e_2$ 
represent the \emph{causality relation}, a 
partial order reflecting the causal relationship
between the atomic events of computation.
The red edges $e_1 \arrowCF e_2$ represent 
the \emph{conflict relation} which is 
a symmetric and irreflexive relation 
encoding mutually exclusive events.
Each particular trace can be extracted from the event structure
as a linearisation of some \emph{configuration}, 
that is a causally-closed and conflict-free 
subset of events.

@@

The programming languages theory and formal semantics research communities 
are moving to increase the usage of \emph{proof assistants} 
like Coq, Agda, Isabelle, Arend, and others,
to complement theoretical studies with their mechanisation,
as this process increases the reliability and reproducibility 
of scientific results.
Yet, to the best of our knowledge, there is little work on 
mechanisation of the theory of event structures.
The present report aims to close the gap.

@@

>>>>

The same information can be encoded in a single 
event structure containing $6$ events in total
(see \cref{fig:intro-es}). 
In the event structure there are two types of edges 
between the events. The grey arrows $e_1 \arrowCA e_2$ 
represent the \emph{causality relation}, a 
partial order reflecting the causal relationship
between the atomic events of computation.
The red edges $e_1 \arrowCF e_2$ represent 
the \emph{conflict relation} which is 
a symmetric and irreflexive relation 
encoding mutually exclusive events.
Each particular trace can be extracted from the event structure
as a linearisation of some \emph{configuration}, 
that is a causally-closed and conflict-free subset of events,  
which additionally should satisfy constraint 
that each read is preceded by a matching write.

@@

The programming languages theory and formal semantics research communities 
are moving to increase the usage of \emph{proof assistants} 
like Coq~\cite{Coq}, Agda~\cite{Agda}, 
Isabelle/HOL~\cite{Nipkow-al:Isabelle2002}, 
Arend~\cite{Arend}, and others,
to complement theoretical studies with their mechanisation,
as this process increases the reliability and reproducibility 
of scientific results.
Yet, to the best of our knowledge, there is little work on 
mechanisation of the theory of event structures.
The present report aims to close the gap.

@@

Our library together with the examples of 
its usage is available online at \url{https://github.com/volodeyka/event-struct}.

### 2021-05-07 papers/SYRCoSE/regmachine-eventstruct.tex
SYRCoSE: minor
*: apply Reviewer #3 suggestions

In this section we present operational semantics
which constructs a prime event structure encoding a 
behaviour of the register machine.

@@

The rules of operational semantics constructing 
the event structure are presented on~\cref{fig:eventstruct-sem}.
The first auxiliary rule \ESAddEventRule adds a new event, sets its 
label, $\lPO$ and $\lRF$ predecessors. 
The \ESIdleRule handles the case when a thread of 
the register machine performs an internal step 
without any side effect. 
It chooses an event $e$ together with 
the thread state $s$ corresponding to it
and performs one step reduction to a new state $s'$.
It then updates the mapping of events to thread states.   
The last two rules \ESStoreRule and \ESLoadRule
correspond to store and load performed by some thread.  
Similarly to \ESIdleRule, an event $e_{\lPO}$ is selected
and one reduction is performed from the corresponding thread state $s$.
Unlike the \ESIdleRule case, however, a new event $e$ is also generated.
In case of \ESLoadRule additionally an event $e_{\lRF}$ is selected,
such that it has a write label matching the read label of the new event.

@@

\begin{theorem}
  The tuple $\tup{E, \caOrd, \primcfRel}$, where $\caOrd$ and $\primcfRel$
  are defined as described above, forms prime event structure.
\end{theorem}

We sketch the proof below.

@@

>>>>

In this section we present operational semantics
which constructs a prime event structure encoding a 
set of possible behaviours of the register machine.

@@

The rules of operational semantics constructing 
the event structure are presented on~\cref{fig:eventstruct-sem}.
The first auxiliary rule \ESAddEventRule~ adds a new event, sets its 
label, $\lPO$ and $\lRF$ predecessors. 
The \ESIdleRule~ handles the case when a thread of 
the register machine performs an internal step 
without any side effect. 
It chooses an event~$e$ together with 
the thread state~$s$ corresponding to it
and performs one step reduction to a new state~$s'$.
It then updates the mapping of events to thread states.   
The last three rules \ESStoreRule, \ESLoadRule, and \ESLoadBotRule~
correspond to store and load performed by some thread.  
Similarly to \ESIdleRule, an event $e_{\lPO}$ is selected
and one reduction is performed from the corresponding thread state~$s$.
Unlike the \ESIdleRule~ case, however, a new event~$e$ is also generated.
In case of \ESLoadRule~ additionally an event~$e_{\lRF}$ is selected,
such that it has a write label matching the read label of the new event.    
The rule \ESLoadBotRule~ corresponds to a case when load 
is performed ``too early'', before any write to the given location is available.

@@

\begin{theorem}
  The tuple $\tup{E, \caOrd, \cfRel}$, where $\caOrd$ and $\cfRel$
  are defined as described above, forms prime event structure.
\end{theorem}

We sketch the proof below\footnote{
One can also find mechanized proof in our Coq development.
}.

@@

For example, a predicate corresponding
to \emph{sequential consistency}~\cite{Lamport:TC79}
requires that the causality order
can be extended to a total order on all events 
of the configuration, such that for each read 
event the last preceding write event 
to the same location has the same value as the read.

### 2021-05-07 papers/SYRCoSE/regmachine.tex
add: finmap + relation-algebra
*: apply Reviewer #3 suggestions

The semantics is given in the form of a
\emph{labelled transition system}~%
\footnote{As we have mentioned, in our \coq development we actually use 
the monadic encoding of the operational semantics. 
The labelled transition system can be derived from this encoding.}.
The state of the system consists of an instruction pointer
and a map from registers to their values, 
as shown in \cref{fig:regmachine-thrdstate}. 
The rules of the semantics are standard 
(see \cref{fig:regmachine-thrdsem}).

>>>>

The semantics is given in the form of a
\emph{labelled transition system}:~%
\footnote{As we have mentioned, in our \coq~development we actually use 
the monadic encoding of the operational semantics. 
The labelled transition system can be derived from this encoding.}
$P \vdash s \thrdstep{l} s'$,
where $P$ is a program, $l$ is a label, $s$ and $s'$ are states of the machine. 
The state of the machine itself consist of an instruction pointer $\ip{i}$
and a map from registers to their values $\regmap$, 
as shown in \cref{fig:regmachine-thrdstate}. 
The rules of the semantics are standard 
(see \cref{fig:regmachine-thrdsem}).
